Estimate of the Neural Net Dimension Using Algebraic Topology and Lie Theory\thanks{The code can be found at: \href{https://github.com/karhunenloeve/NNTopoLie}{https://github.com/karhunenloeve/NNTopoLie}.}

### 2020-06-19 llncs2e/paper12.tex
finish

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{euler}
\usepackage{amsmath,amssymb,amscd,amstext}
\usepackage{xcolor}
\usepackage{pst-node}
\usepackage{tikz,pgf}
\usepackage{tikz-cd} 
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\pgfplotsset{compat=newest}
\usetikzlibrary{spy,decorations.markings,decorations.text,arrows.meta,bending,3d,shapes.arrows,positioning,fit,backgrounds,fadings}
\let\clipbox\relax
\usepackage{slashbox}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{adjustbox}
\usepackage{setspace}
\usepackage{xcolor}
\usepackage{multibib}
\usepackage{filecontents}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{mathtools}
\usepackage{commath}
\usepackage[percent]{overpic}
\usepackage[hidelinks]{hyperref}
\hypersetup{
    pdftitle={Estimate of the Neural Net Dimension Using Algebraic Topology and Lie Theory},    % title
    pdfauthor={Luciano Melodia},     % author
    pdfsubject={Estimate of the Neural Net Dimension},   % subject of the document
    pdfcreator={Luciano Melodia},   % creator of the document
    pdfproducer={Luciano Melodia},  % producer of the document
    pdfkeywords={Embedding dimension} {Parametrization} {Persistent homology} {Neural nets} {Manifold learning}, % list of keywords
    pdfnewwindow=true,
    colorlinks,
	citecolor=black,
	filecolor=black,
	linkcolor=black,
	urlcolor=purple
}
\definecolor{red1}{RGB}{227,226,223}
\definecolor{red2}{RGB}{227,175,188}
\definecolor{red3}{RGB}{238,76,124}
\definecolor{red4}{RGB}{154,23,80}
\definecolor{red5}{RGB}{93,0,30}
\makeatletter
\newcommand{\@chapapp}{\relax}
\newcommand{\xdashrightarrow}[2][]{\ext@arrow 0359\rightarrowfill@@{#1}{#2}}
\newcommand{\xdashleftarrow}[2][]{\ext@arrow 3095\leftarrowfill@@{#1}{#2}}
\newcommand{\xdashleftrightarrow}[2][]{\ext@arrow 3359\leftrightarrowfill@@{#1}{#2}}
\def\rightarrowfill@@{\arrowfill@@\relax\relbar\rightarrow}
\def\leftarrowfill@@{\arrowfill@@\leftarrow\relbar\relax}
\def\leftrightarrowfill@@{\arrowfill@@\leftarrow\relbar\rightarrow}
\def\arrowfill@@#1#2#3#4{%
  $\m@th\thickmuskip0mu\medmuskip\thickmuskip\thinmuskip\thickmuskip
   \relax#4#1
   \xleaders\hbox{$#4#2$}\hfill
   #3$%
}
\newcommand{\ra}[1]{\renewcommand{\arraystretch}{#1}}
\makeatother
\begin{document}
%
\title{Estimate of the Neural Net Dimension Using Algebraic Topology and Lie Theory\thanks{The code can be found at: \href{https://github.com/karhunenloeve/NNTopoLie}{https://github.com/karhunenloeve/NNTopoLie}.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\titlerunning{Estimate of the Neural net Dimension}

@@

\subsubsection*{Embedding dimension of neural nets} Cybenko showed that dense neural nets with one hidden layer can approximate any continuous function with a bounded domain with arbitrarily small error \cite{Cybenko92}. The upper bound of the embedding dimension for ReLU nets was recently determined to be $(n+4)$, where $n$ is the sample dimension \cite{RaghuPKGS17}. Bartlett et al. show an almost tight bound for VC dimension. A neural net with $W$ weights and $L$ layers has $O(WL \log W)$ but $\Omega(WL \log W/L)$ VC dimensions \cite{BartlettHLM19}. Its depth was determined as a parameter depending on the moduli of continuity of the function to be approximated \cite{LinJ18}.

@@

\section{Smooth manifolds, Lie groups and persistent homology}
The assumption that a set of points lies on a manifold is more accurate than the treatment in Euclidean space. A \emph{topological manifold} $M$ is a \emph{Hausdorff} space, which means that at any two points $x,y \in M$ have always open neighborhoods $U_x,U_y \subset M$, so that their intersection is empty. This behaviour gives an intuitive feeling for the position of points in space. The topology $(M, \tau)$ is a set system that describes the structure of a geometric object. Here $M$ itself and $\emptyset$ must be contained in $\tau$, and any union of open sets from $\tau$ and any intersection of $\tau$ must be contained in $\tau$. If this set system has a subset which generates the topology $\tau$ with finitely many unions, then this subset is a \emph{basis} of the topology. We demand of a manifold that the basis is at most countable. In addition, $M$ is locally Euclidean, i.e. each point has a neighborhood which can be mapped homeomorphically to a subset of $\mathbb{R}^n$. The integer $n$ is the dimension of $M$.

\subsection{Smooth manifolds}
A smooth structure is a stronger condition than a topological manifold which can be described by a family of continuous functions. This assumption is justified by the theorem of Stone-Weierstrass which proves that every continuous function can be approximated arbitrarily exactly by a smooth one \cite{stone1948generalized}. Smooth manifolds are described by a family of local coordinate maps $\varphi: U \rightarrow \varphi(U) \subseteq \mathbb{R}^n$ which are homeomorphic to a subset of $\mathbb{R}^{n}$ and cover an open neighborhood $(U_i,\varphi_i)$ in $M$. This family of such charts is called \emph{atlas} on $M$, i.e. $\mathscr{A} = (U_i,\varphi_i)_{i \in I}$ \cite{lee2013smooth}. The functions $x_1, \cdots, x_n: U \rightarrow \mathbb{R}$ are called local coordinates $\varphi(p) = (x_1(p), \cdots, x_n(p))$. If the atlas is maximal in terms of inclusion, then it is a \emph{differentiable structure}. Each atlas for $M$ is included in a maximal atlas. Due to the differentiable structure, the maps in $\mathscr{A}$ are also compatible with all maps of $\mathscr{A}_{\text{max}}$. It is sufficient to describe smooth manifolds by a non maximal atlas.

@@

We apply a theorem from Lie theory, which states that each connected Abelian Lie group $(M,\bullet)$ of dimension $\dim M = n$ is isomorphic to a product space $\mathbb{R}^p \times \mathbb{T}^q$ with $p+q =n$, for a proof we refer to \cite[p.~116]{onishchik1993lie}. The $q$-torus $\mathbb{T}^q$ is a surface of revolution. It moves a curve around an axis of rotation. These axes are given by $1$-spheres, such that the $q$-torus is a product space $\mathbb{T}^q = \mathcal{S}^1_1 \times \cdots \times \mathcal{S}^1_q$. The initial decomposition of connected commutative Lie groups can be further simplified into $M \simeq \mathbb{R}^p \times (\mathcal{S}^1_1 \times \cdots \times \mathcal{S}^1_q)$. Recall, that the $(n-1)$-sphere is given by $\mathcal{S}^{n-1} := \{x \in \mathbb{R}^n \; \vert \; \norm{x}_2 = 1\}$. Thus, $\mathcal{S}^1$ can be embedded into $\mathbb{R}^2$. Next, we derive how many dimensions are at least needed for a suitable embedding. For each $1$-sphere we count two dimensions and for each real line accordingly one. Finally, we have to estimate from data how the decomposition may look like to have the topology of the data manifold.

\subsection{Persistent homology}
Algebraic topology provides a computable tool to study not the topology of a set of points themselfs but an Abelian group attached to them. The core interest in this discipline lies in \emph{homotopy equivalences}, an equivalence class to which objects belong that are continuously deformable into one another. This is much broader than homeomorphism. For two topological spaces $X$ and $Y$ we seek for a function $h: X \times I \rightarrow Y$, which gives the identity at time $h_0(X) = X$ and for $h_1(X) = X$ a mapping into another topological space. If this mapping is continuous with respect to its arguments, it is called \emph{homotopy}. Consider two functions $f,g: I \rightarrow X$ so that $f(1) = g(0)$. Then there is a \emph{composition} of product paths $f \cdot g$, which pass first through $f$ and then through $g$ and which is defined as $f \cdot g (s) = f(2s)$ for $0 \leq s \leq 1/2$ and $g(2s - 1)$, for $1/2 \leq s \leq 1$. We first run $f$ at double speed up to $1/2$ and from $1/2$ to $1$ we run the function $g$ at double speed. In addition, suppose that a family of functions $f: I \rightarrow X$ is given which have the same start and end point $f(0) = f(1) = x_0 \in X$. They intuitively form a \emph{loop}. The set of all homotopy classes $[f]$ of loops $f: I \rightarrow X$ at base point $x_0$ is noted as $\pi_1(X,x_0)$ and is called first homotopy group or fundamental group. The group operation is the product of equivalence classes $[f][g] = [f \cdot g]$. If we do not look at the interval $I$, but at mappings considering the unit cube $I^n$, we obtain the $n$-th homotopy group $\pi_n(X,x_0)$ by analogy. With the help of homotopy groups we study connected components of a topological space for the $0$th group, the loops for the $1$th group, the cavities for the $2$th and so forth. Since homotopy groups are difficult to compute, we resort to an algebraic variant, the \emph{homology groups}.

@@

\input{figure1.tex}
\subsubsection{Simplicial complexes} A \emph{simplicial complex} $K$ is a set of simplices, such that any face of a simplex of $K$ is a simplex of $K$ and the intersection of any two simplices of $K$ is either empty or a common face of both \cite[p.~11]{boissonnat2018geometric}. We require every $(k+1)$-element, with $k < n$ to be a $k$-simplex of $[\sigma] \subset K$. This defines the simplicial complex as a topological space. These properties ensure that simplices are added to the complex in a way, such that they are \emph{glued} edge to edge, point to point, and surface to surface. Thus, every $n$-simplex has $n+1$ distinct vertices and no other $n$-simplex has the same set of vertices. This is a unique combinatorial description of vertices together with a collection of sets $\{[\sigma]_1, \cdots, [\sigma]_m\}$ of $k$-simplices, which are $k+1$ element subsets of $K$ \cite[p.~107]{hatcher2002algebraic}. This set system can be realized geometrically with a homeomorphism into Euclidean space.

\subsubsection{Associated Abelian groups} Algebraically, the simplices can be realized as a system of linear combinations as $\sum_i \lambda_i [\sigma]_{i}$ following Eq. \ref{simplex}. Together with addition as an operation, they form a group structure called $k$th chain group $\langle C_k, +\rangle$ for the corresponding dimension of the $k$-simplices used. The group is commutative or Abelian. The used coefficients $\lambda_i$ induce an orientation, intuitively a direction of the edges, by negative and positive signs, such that $[\sigma] = -[\tau]$, iff $[\sigma] = [\tau]$ and $[\sigma]$ and $[\tau]$ have different orientations. The objects of the chain group are called $k$-chains, for $[\sigma] \in C_k$. The groups are connected by a homomorphism, a mapping which respects the algebraic structure, the boundary operator: $\partial_k: C_k \rightarrow C_{k-1}$, $\partial_k [\sigma] \mapsto \sum_i (-1)^i [v_0,v_1,\cdots,\hat{v}_i, \cdots, v_n]$. The $i$th face is omitted, alternatingly. Using the boundary operator we yield the following sequence of Abelian groups, which is a chain complex cf. Fig. \ref{figure1} a):
\begin{equation}
    0 \xrightarrow[]{\partial_{k+1}} C_k \xrightarrow[]{\partial_{k}} C_{k-1} \xrightarrow[]{\partial_{k-1}} \cdots \xrightarrow[]{\partial_{2}} C_1 \xrightarrow[]{\partial_{1}} C_0 \xrightarrow[]{\partial_{0}} 0.
\end{equation}

@@

\subsubsection{Homological persistence} Examining the homology groups of a set of points gives little information about the structure of a dataset. Instead, we are interested in a parametrization of the simplicial complex as geometric realization in which the homology groups appear and disappear again. For this purpose we consider all possible subcomplexes that form a \emph{filtration} over the point set $X$. We denote $K^{\epsilon} := K^{\epsilon}(X)$. Depending on how we vary the parameter $\epsilon_i$ of the chosen simplicial complex, the following sequence is generated, connected by inclusion and starting with the empty set cf. Fig. \ref{figure1} b):
\begin{align}
    \emptyset &= K^{\epsilon_0} \subset K^{\epsilon_1} \subset \cdots \subset K^{\epsilon_{n+1}}= K,\\
    K^{\epsilon_{i+1}} &= K^{\epsilon_i} \cup \sigma^{\epsilon_{i+1}}, \quad \text{for} \; i \in \{0, \cdots, n-1\}.
\end{align}
The filtration has a discrete realization with a fixed $\epsilon = \epsilon_{i+1}-\epsilon_{i} = \min(x,y)$ for all $x,y \in X$. Through filtration we are able to investigate the homology groups during each step of the parameterization. We record when elements from a homology group appear and when they disappear again. Intuitively speaking, we can see when $k$-dimensional holes appear and disappear in the filtration. We call this process \emph{birth} and \emph{death} of topological features. Recording the \emph{Betti numbers} of the $k$-th homology group along the filtration, we obtain the \emph{$k$-dimensional persistence diagram}, see Fig. \ref{figure1} c)3. The Betti numbers $\beta_k$ are defined by $\text{rank } H_k$. We write $H^{\epsilon_i}_k$ as $k$-th homology group on the simplicial complex $K$ with parametrization $\epsilon_i$. Then $H^{\epsilon_{i}}_{k} \rightarrow H^{\epsilon_{i+1}}_k$ induces a sequence of homomorphisms on the filtration, for a proof we refer to \cite{edelsbrunner2008persistent}:
\begin{align}
0 = H^{\epsilon_{0}}_k \rightarrow H^{\epsilon_{1}}_k \rightarrow \cdots \rightarrow H^{\epsilon_{n}}_k \rightarrow H^{\epsilon_{n+1}}_k = 0.
\end{align}
The image of each homomorphism consists of all $k$-dimensional homology classes which are born in $K^{\epsilon_i}$ or appear before and die after spanning $K^{\epsilon_{i+1}}$. Tracking the Betti numbers on the filtration results into a multiplicity
\begin{equation}
\mu^{\epsilon_{i},\epsilon_{j}}_k = (\beta_k^{\epsilon_{i},\epsilon_{j-1}} - \beta_k^{\epsilon_i,\epsilon_{j}})-(\beta_k^{\epsilon_{i-1},\epsilon_{j-1}}-\beta_k^{\epsilon_{i-1},\epsilon_{j}})
\end{equation}
for the $k$-th homology group and index pairs $(\epsilon_i,\epsilon_{j+1}) \in \overline{\mathbb{R}^2} := \mathbb{R}^2 \cup \infty$ with indices $i \leq j$. The Euclidean space is extended, as the very first connected component on the filtration remains connected. Thus, we assign to it infinite persistence, corresponding to the second coordinate $\epsilon_{j+1}$. The first term counts elements born in $K^{\epsilon_{j-1}}$ and which vanish entering $K^{\epsilon_{j}}$, while the second term counts the representatives of homology classes before $K^{\epsilon_{j}}$ and which vanish at $K^{\epsilon_{j}}$. The $k$-th persistence diagram is then defined as
\begin{equation}
\text{PH}_{k}(X) := \left\{(\epsilon_i, \epsilon_{j+1}) \in \overline{\mathbb{R}^2} \; \bigg\vert \; \mu^{\epsilon_{i},\epsilon_{j+1}}_k = 1 \; \text{for all} \; i,j \in I \right\}.
\end{equation}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{euler}
\usepackage{amsmath,amssymb,amscd,amstext}
\usepackage{xcolor}
\usepackage{pst-node}
\usepackage{tikz,pgf}
\usepackage{tikz-cd} 
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\pgfplotsset{compat=newest}
\usetikzlibrary{spy,decorations.markings,decorations.text,arrows.meta,bending,3d,shapes.arrows,positioning,fit,backgrounds,fadings}
\let\clipbox\relax
\usepackage{slashbox}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{adjustbox}
\usepackage{setspace}
\usepackage{xcolor}
\usepackage{multibib}
\usepackage{filecontents}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{mathtools}
\usepackage{commath}
\usepackage[percent]{overpic}
\usepackage[hidelinks]{hyperref}
\hypersetup{
    pdftitle={Estimate of the Neural Net Dimension Using Algebraic Topology and Lie Theory},    % title
    pdfauthor={Luciano Melodia},     % author
    pdfsubject={Estimate of the Neural Net Dimension},   % subject of the document
    pdfcreator={Luciano Melodia},   % creator of the document
    pdfproducer={Luciano Melodia},  % producer of the document
    pdfkeywords={Embedding dimension} {Parametrization} {Persistent homology} {Neural nets} {Manifold learning}, % list of keywords
    pdfnewwindow=true,
    colorlinks,
	citecolor=black,
	filecolor=black,
	linkcolor=black,
	urlcolor=purple
}
\definecolor{red1}{RGB}{227,226,223}
\definecolor{red2}{RGB}{227,175,188}
\definecolor{red3}{RGB}{238,76,124}
\definecolor{red4}{RGB}{154,23,80}
\definecolor{red5}{RGB}{93,0,30}
\makeatletter
\newcommand{\@chapapp}{\relax}
\newcommand{\xdashrightarrow}[2][]{\ext@arrow 0359\rightarrowfill@@{#1}{#2}}
\newcommand{\xdashleftarrow}[2][]{\ext@arrow 3095\leftarrowfill@@{#1}{#2}}
\newcommand{\xdashleftrightarrow}[2][]{\ext@arrow 3359\leftrightarrowfill@@{#1}{#2}}
\def\rightarrowfill@@{\arrowfill@@\relax\relbar\rightarrow}
\def\leftarrowfill@@{\arrowfill@@\leftarrow\relbar\relax}
\def\leftrightarrowfill@@{\arrowfill@@\leftarrow\relbar\rightarrow}
\def\arrowfill@@#1#2#3#4{%
  $\m@th\thickmuskip0mu\medmuskip\thickmuskip\thinmuskip\thickmuskip
   \relax#4#1
   \xleaders\hbox{$#4#2$}\hfill
   #3$%
}
\newcommand{\ra}[1]{\renewcommand{\arraystretch}{#1}}
\makeatother
\begin{document}
%
\title{Estimate of the Neural Net Dimension Using Algebraic Topology and Lie Theory\thanks{The code can be found at: \href{https://github.com/karhunenloeve/NNTopoLie}{https://github.com/karhunenloeve/NNTopoLie}.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\titlerunning{Estimate of the Neural Net Dimension}

@@

\subsubsection*{Embedding dimension of neural nets} Cybenko showed that dense neural nets with one hidden layer can approximate any continuous function with a bounded domain with arbitrarily small error \cite{Cybenko92}. The upper bound of the embedding dimension for ReLU nets was recently determined to be $(n+4)$, where $n$ is the sample dimension \cite{RaghuPKGS17}. Bartlett et al. show an almost tight bound for VC dimensions. A neural net with $W$ weights and $L$ layers has $O(WL \log W)$ but $\Omega(WL \log W/L)$ VC dimensions \cite{BartlettHLM19}. Its depth was determined as a parameter depending on the moduli of continuity of the function to be approximated \cite{LinJ18}.

@@

\section{Smooth manifolds, Lie groups and persistent homology}
The assumption that a set of points lies on a manifold is more accurate than the treatment in Euclidean space. A \emph{topological manifold} $M$ is a \emph{Hausdorff} space, which means that any two points $x,y \in M$ have always open neighborhoods $U_x,U_y \subset M$, so that their intersection is empty. This behaviour gives an intuitive feeling for the position of points in space. The topology $(M, \tau)$ is a set system that describes the structure of a geometric object. Here $M$ itself and $\emptyset$ must be contained in $\tau$, and any union of open sets from $\tau$ and any intersection of $\tau$ must be contained in $\tau$. If this set system has a subset which generates the topology $\tau$ with finitely many unions, then this subset is a \emph{basis} of the topology. We demand of a manifold that the basis is at most countable. In addition, $M$ is locally Euclidean, i.e. each point has a neighborhood which can be mapped homeomorphically to a subset of $\mathbb{R}^n$. The integer $n$ is the dimension of $M$.

\subsection{Smooth manifolds}
A smooth structure is a stronger condition than a topological manifold which can be described by a family of continuous functions. This assumption is justified by the theorem of Stone-Weierstrass which proves that every continuous function can be approximated arbitrarily exactly by a smooth one \cite{stone1948generalized}. Smooth manifolds are described by a family of local coordinate maps $\varphi: U \rightarrow \varphi(U) \subseteq \mathbb{R}^n$ which are homeomorphic to a subset of $\mathbb{R}^{n}$ and cover an open neighborhood $(U_i,\varphi_i)$ in $M$. A family of charts, which covers $M$, is called \emph{atlas} on $M$, i.e. $\mathscr{A} = (U_i,\varphi_i)_{i \in I}$ \cite{lee2013smooth}. The functions $x_1, \cdots, x_n: U \rightarrow \mathbb{R}$ are called local coordinates $\varphi(p) = (x_1(p), \cdots, x_n(p))$. If the atlas is maximal in terms of inclusion, then it is a \emph{differentiable structure}. Each atlas for $M$ is included in a maximal atlas. Due to the differentiable structure, the maps in $\mathscr{A}$ are also compatible with all maps of $\mathscr{A}_{\text{max}}$. It is sufficient to describe smooth manifolds by a non maximal atlas.

@@

We apply a theorem from Lie theory, which states that each connected Abelian Lie group $(M,\bullet)$ of dimension $\dim M = n$ is isomorphic to a product space $\mathbb{R}^p \times \mathbb{T}^q$ with $p+q =n$, for a proof we refer to \cite[p.~116]{onishchik1993lie}. The $q$-torus $\mathbb{T}^q$ is a surface of revolution. It moves a curve around an axis of rotation. These axes are given by $1$-spheres, such that the $q$-torus is a product space $\mathbb{T}^q = \mathcal{S}^1_1 \times \cdots \times \mathcal{S}^1_q$. The initial decomposition of connected commutative Lie groups can be further simplified into $M \simeq \mathbb{R}^p \times (\mathcal{S}^1_1 \times \cdots \times \mathcal{S}^1_q)$. Recall, that the $(n-1)$-sphere is given by $\mathcal{S}^{n-1} := \{x \in \mathbb{R}^n \; \vert \; \norm{x}_2 = 1\}$. Thus, $\mathcal{S}^1$ can be embedded into $\mathbb{R}^2$. Next, we derive how many dimensions are at least needed for a suitable embedding. For each $1$-sphere we count two dimensions and for each real line accordingly one. Finally, we have to estimate from data how the decomposition may look like to yield the topology of the data manifold.

\subsection{Persistent homology}
Algebraic topology provides a computable tool to study not the topology of a set of points themselfs but an Abelian group attached to them. The core interest in this discipline lies in \emph{homotopy equivalences}, an equivalence class to which objects belong that are continuously deformable into one another. This is much broader than homeomorphism. For two topological spaces $X$ and $Y$ we seek for a function $h: X \times I \rightarrow Y$, which gives the identity at time $h_0(X) = X$ and for $h_1(X) = Y$ a mapping into another topological space. If this mapping is continuous with respect to its arguments, it is called \emph{homotopy}. Consider two functions $f,g: I \rightarrow X$ so that $f(1) = g(0)$. Then there is a \emph{composition} of product paths $f \cdot g$, which pass first through $f$ and then through $g$ and which is defined as $f \cdot g (s) = f(2s)$ for $0 \leq s \leq 1/2$ and $g(2s - 1)$, for $1/2 \leq s \leq 1$. We first run $f$ at double speed up to $1/2$ and from $1/2$ to $1$ we run the function $g$ at double speed. In addition, suppose that a family of functions $f: I \rightarrow X$ is given which have the same start and end point $f(0) = f(1) = x_0 \in X$. They intuitively form a \emph{loop}. The set of all homotopy classes $[f]$ of loops $f: I \rightarrow X$ at base point $x_0$ is noted as $\pi_1(X,x_0)$ and is called first homotopy group or fundamental group. The group operation is the product of equivalence classes $[f][g] = [f \cdot g]$. If we do not look at the interval $I$, but at mappings considering the unit cube $I^n$, we obtain the $n$-th homotopy group $\pi_n(X,x_0)$ by analogy. With the help of homotopy groups we study connected components of a topological space for the $0$th group, the loops for the $1$th group, the cavities for the $2$th and so forth. Since homotopy groups are difficult to compute, we resort to an algebraic variant, the \emph{homology groups}.

@@

\input{figure1.tex}
\subsubsection{Simplicial complexes} A \emph{simplicial complex} $K$ is a set of simplices, such that any face of a simplex of $K$ is a simplex of $K$ and the intersection of any two simplices of $K$ is either empty or a common face of both \cite[p.~11]{boissonnat2018geometric}. We require every $(k+1)$-element, with $k < n$ to be a $k$-simplex of $[\sigma] \subset K$. This defines the simplicial complex as a topological space. These properties ensure that simplices are added to the complex in a way, such that they are \emph{glued} edge to edge, point to point, and surface to surface. Thus, every $n$-simplex has $n+1$ distinct vertices and no other $n$-simplex has the same set of vertices. This is a unique combinatorial description of vertices together with a collection of sets $\{[\sigma]_1, \cdots, [\sigma]_m\}$ of $k$-simplices, which are $k+1$ element subsets of $K$ \cite[p.~107]{hatcher2002algebraic}. This set system can be realized geometrically with an injective map into Euclidean space.

\subsubsection{Associated Abelian groups} Algebraically, the simplices can be realized as a system of linear combinations as $\sum_i \lambda_i [\sigma]_{i}$ following Eq. \ref{simplex}. Together with addition as an operation, they form a group structure called $k$th chain group $\langle C_k, +\rangle$ for the corresponding dimension of the $k$-simplices used. The group is commutative or Abelian. The used coefficients $\lambda_i$ induce an orientation, intuitively a direction of the edges, by negative and positive signs, such that $[\sigma] = -[\tau]$, iff $[\sigma] \sim [\tau]$ and $[\sigma]$ and $[\tau]$ have different orientations. The objects of the chain group are called $k$-chains, for $[\sigma] \in C_k$. The groups are connected by a homomorphism, a mapping which respects the algebraic structure, the boundary operator: $\partial_k: C_k \rightarrow C_{k-1}$, $\partial_k [\sigma] \mapsto \sum_i (-1)^i [v_0,v_1,\cdots,\hat{v}_i, \cdots, v_n]$. The $i$th face is omitted, alternatingly. Using the boundary operator we yield the following sequence of Abelian groups, which is a chain complex cf. Fig. \ref{figure1} a):
\begin{equation}
    0 \xrightarrow[]{\partial_{k+1}} C_k \xrightarrow[]{\partial_{k}} C_{k-1} \xrightarrow[]{\partial_{k-1}} \cdots \xrightarrow[]{\partial_{2}} C_1 \xrightarrow[]{\partial_{1}} C_0 \xrightarrow[]{\partial_{0}} 0.
\end{equation}

@@

\subsubsection{Homological persistence} Examining the homology groups of a set of points gives little information about the structure of a dataset. Instead, we are interested in a parametrization of the simplicial complex as geometric realization in which the homology groups appear and disappear again. For this purpose we consider all possible subcomplexes that form a \emph{filtration} over the point set $X$. We denote $K^{\epsilon} := K^{\epsilon}(X)$. Depending on how we vary the parameter $\epsilon_i$ of the chosen simplicial complex, the following sequence is generated, connected by inclusion and starting with the empty set cf. Fig. \ref{figure1} b):
\begin{align}
    \emptyset &= K^{\epsilon_0} \subset K^{\epsilon_1} \subset \cdots \subset K^{\epsilon_{n+1}}= K,\\
    K^{\epsilon_{i+1}} &= K^{\epsilon_i} \cup [\sigma]^{\epsilon_{i+1}}, \quad \text{for} \; i \in \{0, \cdots, n-1\}.
\end{align}
The filtration has a discrete realization with a fixed $\epsilon = \epsilon_{i+1}-\epsilon_{i} = \min(x,y)$ for all $x,y \in X$. Through filtration we are able to investigate the homology groups during each step of the parameterization. We record when elements from a homology group appear and when they disappear again. Intuitively speaking, we can see when $k$-dimensional holes appear and disappear in the filtration. We call this process \emph{birth} and \emph{death} of topological features. Recording the \emph{Betti numbers} of the $k$-th homology group along the filtration, we obtain the \emph{$k$-dimensional persistence diagram}, see Fig. \ref{figure1} c)3. The Betti numbers $\beta_k$ are defined by $\text{rank } H_k$. We write $H^{\epsilon_i}_k$ as $k$-th homology group on the simplicial complex $K$ with parametrization $\epsilon_i$. Then $H^{\epsilon_{i}}_{k} \rightarrow H^{\epsilon_{i+1}}_k$ induces a sequence of homomorphisms on the filtration, for a proof we refer to \cite{edelsbrunner2008persistent}:
\begin{align}
0 = H^{\epsilon_{0}}_k \rightarrow H^{\epsilon_{1}}_k \rightarrow \cdots \rightarrow H^{\epsilon_{n}}_k \rightarrow H^{\epsilon_{n+1}}_k = 0.
\end{align}
The image of each homomorphism consists of all $k$-dimensional homology classes which are born in $K^{\epsilon_i}$ or appear before and die after spanning $K^{\epsilon_{i+1}}$. Tracking the Betti numbers on the filtration results into a multiplicity
\begin{equation}
\mu^{\epsilon_{i},\epsilon_{j}}_k = (\beta_k^{\epsilon_{i},\epsilon_{j-1}} - \beta_k^{\epsilon_i,\epsilon_{j}})-(\beta_k^{\epsilon_{i-1},\epsilon_{j-1}}-\beta_k^{\epsilon_{i-1},\epsilon_{j}}),
\end{equation}
for the $k$-th homology group and index pairs $(\epsilon_i,\epsilon_{j+1}) \in \overline{\mathbb{R}^2} := \mathbb{R}^2 \cup \infty$ with indices $i \leq j$. The Euclidean space is extended, as the very first connected component on the filtration remains connected. Thus, we assign to it infinite persistence, corresponding to the second coordinate $\epsilon_{j+1}$. The first term counts elements born in $K^{\epsilon_{j-1}}$ and which vanish entering $K^{\epsilon_{j}}$, while the second term counts the representatives of homology classes before $K^{\epsilon_{j}}$ and which vanish at $K^{\epsilon_{j}}$. The $k$-th persistence diagram is then defined as
\begin{equation}
\text{PH}_{k}(X) := \left\{(\epsilon_i, \epsilon_{j+1}) \in \overline{\mathbb{R}^2} \; \bigg\vert \; \mu^{\epsilon_{i},\epsilon_{j+1}}_k = 1 \; \text{for all} \; i,j \in I \right\}.
\end{equation}
